Simulating High-Dimensional Multivariate Data

### 2020-10-29 010-background.Rmd
bigsimr R package rev3

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none'}
if ( !file.exists( 'data/brca99.rds' ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-probs)*100% highest expressing genes for illustration
    myProb <- 0.99
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, 'data/brca99.rds')
}
```

@@

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
brca <- readRDS(file = './data/brca99.rds')
set.seed(10192020)
exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
saveRDS(exampleGenes, file = './data/exampleGenes.rds')
kable(round(head( brca[, exampleGenes], n = 5)),
      booktabs = TRUE,
      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
)
## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

>>>>

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none'}
if ( !file.exists( 'data/brca99.rds' ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-myProb)*100% highest expressing genes for illustration
    myProb <- 0.99
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, 'data/brca99.rds')
}
myProb <- 0.99
brca <- readRDS(file = './data/brca99.rds')
d <- ncol(brca)
```

@@

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
set.seed(10192020)
exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
saveRDS(exampleGenes, file = './data/exampleGenes.rds')
kable(round(head( brca[, exampleGenes], n = 5)),
      booktabs = TRUE,
      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
)
## GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

### 2020-10-29 030-bigsimr-package.Rmd
bigsimr R package rev3

*Specifying marginal distributions*.
Though we don't know the true distribution of ozone levels or air temperatures, we'll assume that both are log-normally distributed since the values are positive and skewed.
We'll use the well-known maximum likelihood estimators for the lognormal parameters:

```{r ch030-aq-temp-pars, echo=FALSE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

@@

df %>% 
  select(Temp, Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

@@

```{r ch030-margins-alist, echo=TRUE}
margins <- alist(
  qlnorm(meanlog = 0.74, sdlog = 0.016),
  qlnorm(meanlog = 3.42, sdlog = 0.743)
)
```

@@

```{r ch030-margins-mlist, echo=TRUE}
margins <- mlist(
  qlnorm(meanlog = mle_mean(df$Temp), sdlog = mle_sd(df$Temp)),
  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
)
margins
```

@@

Since our estimated Spearm correlation $\hat{ \rho_S }$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).

@@

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimick the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Temperature", y = "Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

>>>>

*Specifying marginal distributions*.
Of course, we don't know the true distribution of ozone levels or air temperatures.
Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally ditsributed and `Ozone` is log-normally distributed since its values are positive and skewed.
We'll use the well-known unbiased estimators for normal distribution and maximum likelihood estimators for the lognormal parameters:

```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

@@

df %>% 
  select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

@@

```{r ch030-margins-alist, echo=TRUE}
margins <- alist(
    qnorm(mean = 77.871, sd = 9.4855),
    qlnorm(meanlog = 3.419, sdlog = 0.7426),
)
```

@@

```{r ch030-margins-mlist, echo=TRUE}
margins <- mlist(
  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
)
margins
```

@@

Since our estimated Spearm correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).

@@

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimick the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))
